A multi-scale story of the diffusion of a new technology: the web

### 2024-09-12 paper/submission1/tranos2023.tex
postpresentation2

\affiliation[1]{organization={University of Bristol and The Alan Turing
Institute},postcode={UK},postcodesep={}}

@@

\section{Introduction}\label{sec1}

Geographers were always interested in how new technologies and
innovations diffuse across space and time and, importantly, how such
spatio-temporal processes can be modeled. After all, diffusion together
with invention and innovation are considered the pillars of
technological change \citep{das2022diffusion}. The seminal contribution
of \citet{hagerstrand1968innovation} is illustrative of this early
interest. However, the torch of exploring and modelling such processes
had been passed to other disciplines such as economics, business studies
and sociology well before the `cultural turn' of economic geography
\citep{perkins2005international}. A potential explanation of the lack of
geographical studies exploring the diffusion of new and, more
specifically for this paper, digital technologies across both space and
time can be attributed to the scarcity of relevant and granular enough
data. As \citet{zook2022mapping} highlight, digital activities are
hardly ever captured in official data.

This paper offers such a contribution: a geographical study illustrating
how a new technology that is the web diffused over space and time in the
UK at a high level of spatial granularity during the \(1996-2012\)
period. It does so by employing a novel source of big data which
captures the active engagement with web technologies during that period.
By addressing this empirical question this paper exemplifies how the
combination of data sources which escape the traditional social science
domain and adequate research methods can offer new lenses to
geographical research regarding the understanding of technological
diffusion.

\textbf{SAY SOMETHING ABOUT SPATIAL ANALYTICS, DIST PAPER BY ARP, NOT
SPECIFIC EXPLANATORY VARIABLES, INTANGIBLE}

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
regarding the diffusion of related \emph{future} technologies. As per
\citet{leibowicz2016representing}, historical studies agree on the fact
that ``technologies diffuse at different times, at different rates, and
to different extents in different places, and can be significantly
influenced by policies'' \citep{victor1993}. \citet{meade2021modelling}
highlight that a variety of actors have a direct interest in gaining
such knowledge including network equipment suppliers; network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places \citep{leibowicz2016representing}. Despite the spatial
heterogeneity of such diffusion mechanisms and the policy relevance,
there are very limited attempts in the literature to analyse the
diffusion of new digital technologies at a detailed geographical level.

Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, \citet{rogers2010diffusion} identifies early adopter of new
technologies as `knowledgeable risk takers' and \citet{griliches1957} as
`profit maximisers' \citep{ding2010modeling}. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, \citet{perkins2011internet} attribute
diffusion to two processes: (i) epidemic-like mechanisms, which are
governed by distance, proximity and social interactions, and (ii) by
economic mechanisms as new innovations are adopted by users as they
become more profitable, valueable and useful.

This paper focuses on the diffusion of the web as new technology during
the \(1996-2012\) period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.\footnote{\url{https://www.theguardian.com/global/2015/mar/22/web-browser-came-back-haunt-microsoft}}
Also, only \(9\) per cent of UK's households had access to the internet
in \(1998\) \citep{ons2018}, the web included mostly static webpages,
there were no social media and web browsing was happening exclusively
from desktop PCs as there were no smartphones \citep{tranosuk}. Hence,
it is fair to say that the study period captures the very early stages
of the diffusion on a new technology that is the web as well as its
maturity. The former is a key point in the lifecycle of a new
technology. Firstly, during this period new technologies are expensive,
crude and imperfect \citep{rosenberg1994exploring, wilson201281}. A
simple comparison between Web 1.0 and Web 3.0 applications clearly
illustrates this \citep{tranos2020social}; for instance a static website
compared with a platform like \texttt{github}, which enables cooperation
between users and the creation of new information, meaning, and
knowledge \citep{faraj2016special, barassi2012does}. During this period
the performance of a new technology is the main attraction and not the
cost to access and use it \citep{wilson2011lessons}. There is a broader
theoretical discussion in the literature regarding the motivation behind
early adoption. As summarised by \citet{perkins2005international}, on
the one hand, epidemic models highlight the role of interpersonal
contacts as a way for new technologies to diffuse. On the other hand,
economic models underline the importance of heterogeneity. Different
firms have different structures and business plans, which define the
potential economic returns of the adoption of a new technology and,
therefore, the choice to adopt a new technology becomes an individual
option. From a broader and evolutionary perspective, initial conditions
are essential for the creation and evolution of path-dependent
technological development trajectories
\citep{neffke2011regions, simmie2014new}. This argument is even more
relevant when the focus is on digital technologies because of the
commonly found lag between investment and economic returns as reflected
in the Solow paradox
\citep{acemoglu2014return, brynjolfsson2018artificial}.

Importantly, the data used here depicts the \emph{active} engagement
with the web in the UK as it contains all commercial websites that (i)
are part of the UK's relevant second level domain (SLD, .co.uk), (ii)
have been archived by the Internet Archive\footnote{\href{See\%20https://archive.org/}{https://archive.org/}.},
and (iii) include a mention to at least one valid UK postcode in the web
text. The act of creating a website is understood here as active
engagement with the web vis-à-vis the more passive act of browsing the
web or having an internet connection \citep{tranosuk}. Previous studies
have focused mostly on more passive notions of engaging with digital
technologies such as internet adoption and internet speeds
\citep[e.g.][]{blank2018local, riddlesden2014broadband}. More details
about the data and the data generation process can be found in Section
\hyperref[sec3]{XX}.

\citet{grubler1990rise} Later Hagerstrand conceptualized physical
``barrier'' effects like lakes or uninhabited areas, which, in addition
to distance, act as further retarding effects on diffusion. These are
formalized in the form of ``zero'' or ``half'' contact multiplicators on
the (distance decaying) message flows.

\citet{grubler1990rise} With respect to the formalization of the
communication flows Hagerstrand defines a ``mean information field''
(MIF), in which the probability of communication is a negative function
of distance between individuals

\citet{wilson201281} Logistic growth describes an initial period of
gradual diffusion as a technology is introduced as a new commercial
application, moving then through a rapid, exponential growth phase,
before slowing and eventually saturating \citep{grubler1999dynamics}.
The substitution of incumbent technologies by new competitors leads to
subsequent decline and eventual obsolescence.

\section{Literature review}\label{sec2}

Geographical diffusion is a synthesis of different processes. On the one
hand, we can identify purely spatial or, in other words, contagious
processes. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure \citep{hivner2003facilitating}. On
the other hand, we can identify hierarchical processes. Instead of
horizontal distance-based diffusion mechanisms, the top-down hierarchy
of urban systems drives diffusion. In reality, the synthesis of these
two processes represents how new technologies diffuse over space and
time \citep{morrill2020spatial}.

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled `Innovation Diffusion as a Spatial Process'
\citep{hagerstrand1968innovation}. Hägerstrand was the first one to
identify diffusion as a geographical process. The starting point was the
idea that diffusion is based on passing information through social
networks, which themselves tend to be defined by geography. Hence, he
identified the `neighbourhood' effect of how information, and
consequently, innovation diffuse. He used agricultural innovations to
test and model his ideas using Monte Carlo simulations. Hägerstrand also
incorporated the role of hierarchy and how some phenomena maybe firstly
adopted in larger cities and then diffuse to second tier ones. This is a
sequential instead of a simultaneous process, which resembles the
`lead-lag' spatial acceleration effect in market research
\citep{bento2018time, PERES201091}. Hägerstrand is more widely known
though for highlighting the role time plays in the diffusion of
innovations: an early-pioneering period, a middle fast accelerating
period and a final saturation one \citep{morrill2020spatial}.

The temporal dimension was further explored by Everett Rogers and his
seminal work on `Diffusion of Innovations' \citep{rogers2010diffusion}.
Rogers being a sociologist, his work focused not on the diffusion of
innovations over space and time, but instead on the adoption of new
technologies and innovations by individuals and the individual
mechanisms that drive the decisions behind adoption. He identified five
groups of individuals regarding their adoption speed: innovators, early
adopters, early majority, late majority and laggards. The key mechanism
of diffusion and adoption is communication and how knowledge is
transferred within a social system. Therefore, all approaches agree on
the S-shaped diffusion and cumulative adoption pattern
\citep{grubler1990rise}.

Schmidt's Law empirically illustrates a similar pattern. \emph{Core} and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed \citep{grubler1990rise}. This is
where the first adopters tend to be based. Then, technologies spread to
the \emph{rim} and eventually to the \emph{periphery}. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions \citep{leibowicz2016representing}. \citet{grubler1990rise}
effectively summarises the three key characteristics of the spatial
diffusion process: (i) the cumulative level of adoption follows an
S-shaped pattern just like purely temporal models; (ii) diffusion is
shaped by a hierarchy effect in a form of a centrifugal force: from core
to periphery; and (iii) diffusion is also shaped by distance and a
neighbourhood effect and contaminate nearby locations. These are the
three mechanisms that the empirical analysis in Section
\phantomsection\label{sec4}{4} investigates.

The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted here.
\citet{beardsell1999spatial} studied the evolution of the computer
industry in \(317\) US metro areas during the \(1977-1992\) period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, \citet{bednarz2020pulled} focused on wind turbines and modelled
their spatial diffusion across \(402\) German regions during
\(1970-2015\). Their key finding is that local demand than local supply
was the main driving factor.

At a global scale \citet{perkins2005international} explored whther the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
\citet{perkins2011internet} explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones was
shaped by similar socioeconomic factors as the internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, \citet{ding2010modeling} modelled the spatial diffusion
of mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network,
\citet{lengyel2020role} analysed its adoption and the churn at a very
granular spatial level. Their results are in agreement with early
theoretical and empirical contributions reviewed here: assortativity,
urban scaling and distance are the key drivers of spatial diffusion. At
a global scale \citet{PAPAGIANNIDIS2015308} modelled the diffusion of
different web technologies technologies and practices. Interestingly,
they did so by using similar, but less extensive data as the one used
here. Their analysis illustrated how the diffusion of different web
technologies and practices follow an S-shaped pattern as well as the
different diffusion rates of the different technologies and practices.

All in all, \ldots{}

\section{Materials and Methods}\label{sec3}

\textbf{ADD FAMILY OF WEB TECHNOLOGIES}

To capture the diffusion of web technologies, a website density metric
is developed for two different geographical scales: the Local Authority
Districts (LAD) and the Output Areas (OA). The former is an
administrative unit and there are
\textbackslash textcolor\{red\}c.~380\textbackslash textcolor\{black\}
such units in the UK. The latter is a census-based geographical unit,
which is very small as there are c.~230,000 of them in the UK. This
methodological choice will allow the mapping of the diffusion of web
technologies and the assessment of the diffusion mechanisms at these two
very different spatial scales.

The population of websites is calculated using data from the Internet
Archive \footnote{\href{See\%20https://archive.org/}{https://archive.org/}.}
and, specifically, the JISC UK Web Domain Dataset \citep{ukwebarchive}.
The Internet Archive is one of the most complete and oldest archive of
webpages in the world operating since 1996
\citep{ainsworth2011much, holzmann2016dawn}. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage its
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK ccTLD (.uk) from the
1996--2012 period. In essence, this is a long list of 2.5 billion URLs
of archived webpages including also the archival timestamp.

Instead of using the whole .uk country code top-level domain (ccTLD),
this paper focuses on its commercial subset, the .co.uk SLD. This choice
decrease the heterogeneity of the web data as such commercial websites
have rather specific aims: they are used to diffuse information, support
online transactions and share opinions
\citep{THELWALL2000441, blazquez2018big}. Although a UK company can
adopt a generic TLD such as.com and such cases escape the data used
here, such omissions should not affect our results given the popularity
of the .uk ccTLD \citep{tranosuk}: UK consumers prefer to visit a .uk
website when they are searching for products or services \citep{hope};
and anecdotal evidence indicates that during the first half of 2000,
three .co.uk domains were registered every minute \citep{oecd_coms}.
Importantly, previous studies illustrated that .co.uk is the most
popular UK SLD \citep{tranosuk}.

The text from these webpages was scanned using regular expressions
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode \citep{BL2013geo}. This
information allows the geolocation of the data and the creation of the
LAD and OA counts.

Before that, the data cleaning process included a data aggregation step,
through which the webpages were aggregated to the relevant websites.
Based on the following example, all three webpages listed below are part
of the same website (\url{http://www.website.co.uk}) and at the end only
websites and not the nested webpages were considered as otherwise the
metric would have been biased towards large websites.

\begin{itemize}
\item
  \url{http://www.website.co.uk/webpage_a} B152TT
\item
  \url{http://www.website.co.uk/webpage_b} BS81TH
\item
  \url{http://www.website.co.uk/webpage_c} B152TT
\end{itemize}

\noindent What is challenging is that this aggregation approach, which
has been used elsewhere \citep{tranosuk, shoreditch} may lead to
websites with multiple postcodes. As per the above example,
(\url{http://www.website.co.uk}) includes two unique postcodes: B152TT
and BS81SS. The distribution of postcodes for 2000 is presented in Table
\ref{f2000}, which clearly illustrates the wide range. On one hand, we
have websites anchored to a unique location (72\% of all the
reconstructed websites), which may represent a small company with a
single trading location. On the other hand, we have websites with
thousands of different postcodes. Considering the timeperiod of the
analysis, such cases can represent directories which used to be popular
in the pre-search engines early times of internet
\citep{tranosuk}.\footnote{See Figure A1 in \textbf{Appendix A in the
  supplemental data online for examples}.}

The analysis presented here is based on two subsets of these data.
Firstly, the analysis is based on websites which only contain only one
unique postcode and, therefore, the geolocation process might suffer
from less noise. As a robustness check, the analysis is replicated the
analysis for an extended subset of websites, which include up to ten
unique postcodes. These websites are geolocated by equally attaching
them to multiple locations. This extended sample includes 94 percent of
all the archived websites in 2000.

@@

To create the web density metric, the yearly website counts at the LAD
level are standardised by the number of firms in LAD to avoid biases
associated with LADs hosting a large number of firms. Given that there
is no such statistic for the OA, the actual OA level counts are used.
However, because of the the rather consistent spatial definition of OA
(they host 40-250 households),\footnote{\url{https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies}}
website counts in OA are interpreted as a density metric too.

This website metric\ldots{}

\citet{wilson201281} growth function description in p.~86

from R: Asym/(1+exp((xmid-input)/scal)) using the terms from Wilson k/(1
+ e\^{}(t\_0 - t)/scal)

Wilson: k(1 + e\^{}-b(t - t\_0)) So, scal = - 1/b

The curve is symmetric* around t\_0

\textbf{minimum R2 = 95\%}, see also \citet{grubler1990rise}

The literature usually uses the saturation level as the asymptote. I am
using the total number of websites as we cannot compute a rate.

>>>>

\affiliation[1]{organization={University of
Bristol},postcode={UK},postcodesep={}}

@@

\section{Introduction}\label{sec-introduction}

Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
\citet{hagerstrand1968innovation}, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance \citet{ormrod1990} and \citet{iso2005}. However, the torch
of exploring and modelling such diffusion processes had been passed to
other disciplines such as economics, business studies and sociology well
before the `cultural turn' of economic geography
\citep{perkins2005international}. A potential explanation of the recent
lack of geographical studies exploring the diffusion of new and, more
specifically for this paper, digital technologies across both space and
time can be attributed to the scarcity of relevant and granular enough
data, a problem also highlighted by \citet{iso2005} and
\citet{kemeny2011international}. As \citet{zook2022mapping} pointed,
digital activities are hardly ever captured in official data.

This paper offers such a contribution: a geographical study illustrating
how a new technology that is the Web diffused over space and time in the
UK at a high level of spatial granularity during the 1996-2012 period.
It does so by employing a novel source of big data and machine learning
algorithms to model the \emph{active} engagement with Web during that
period and tests how different spatial and temporal mechanisms shaped
its diffusion. This paper exemplifies how the combination of data
sources which escape the traditional social science domain and adequate
research methods can offer new lenses to geographical research regarding
the understanding of technological diffusion.

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
about the diffusion of related \emph{future} technologies. As per
\citet{leibowicz2016representing}, historical studies agree that
technologies diffuse differently in terms of times, rates, and
geographies and can be driven by related policies \citep{victor1993}.
\citet{meade2021modelling} highlight that a variety of actors have a
direct interest in gaining such knowledge including network equipment
suppliers, network operators, regulatory and local authorities. These
processes and their effects vary a lot across scales: although the
diffusion of a new technology might not be optimal at a local level, it
might be beneficial from a global perspective as it could lead to faster
diffusion to less advantaged places \citep{leibowicz2016representing}.
Despite the spatial heterogeneity of such diffusion mechanisms and the
policy relevance, there is a scarcity of empirical studies analysing the
diffusion of new digital technologies at a detailed geographical level.

This gap becomes even more prominent if we consider the importance of
technology and its adoption in economic geography and macroeconomic
growth theories. Simply put, differences in technological adoption and
sophistication may lead to uneven development as new technologies can
increase productivity and allow firms to introduce new products
\citep{solow1957technical, aghion1990model, kemeny2011international}.
Linked to the subject of this paper, the Web -- and the active
engagement with this technology that is creating instead of just
browsing a website -- can be considered as the underpinning technology
for the transformative process known as the \emph{digital service
economy} \citep{capello2024nexus}. After all, diffusion together with
invention and innovation are considered the pillars of technological
change \citep{das2022diffusion}. Findings from previous empirical work
illustrated that early engagement with digital technologies such as the
Web led to longstanding positive productivity gains for these regions
that hosted early adopters \citep{tranosuk}.

Another line of motivation has to do with the intangible nature of the
Web and the debate about the footloose nature of digital technologies,
which contrary to several earlier general purpose technologies are light
in weight \citep{coyle1999weightless}. Following the seminal work of
\citet{gaspar1998information}, which demonstrated that technological
improvements in telecommunications lead to increased demand for
face-to-face communications and, consequently, enhance the importance of
cities, empirical studies from economic geography and urban economics
explored whether digital technologies substitute or complement
agglomeration externalities and spatial structure
\citep{kolko_death_2000, ioannides2008effect, tranos2021ubiquitous}.
With the exception of \citet{sinai2004geography}, who focused, among
other things, on web content and demonstrated that more such content is
available in larger markets, this strand of literature has been mostly
concerned with the spatiality of the internet's hard and tangible
infrastructure and shied away from exploring the spatial dimensions of
one of the most distinct software layers of the internet infrastructure,
the Web.

Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, \citet{rogers2010diffusion} identifies early adopter of new
technologies as `knowledgeable risk takers' and \citet{griliches1957} as
`profit maximisers' \citep{ding2010modeling}. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, \citet{perkins2011internet} attribute
diffusion to two processes: (i) epidemic-like mechanisms, which are
governed by distance, proximity and social interactions, and (ii)
economic mechanisms as new innovations are adopted by users as they
become more profitable, valueable and useful.

This paper focuses on the diffusion of the Web as new technology during
the 1996-2012 period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.\footnote{\url{https://www.theguardian.com/global/2015/mar/22/web-browser-came-back-haunt-microsoft}}
Also, only 9 per cent of UK's households had access to the internet in
1998 \citep{ons2018}, the web included mostly static webpages, there
were no social media and web browsing involved exclusively a desktop PCs
as there were no smartphones \citep{tranosuk}. Hence, it is fair to say
that the study period captures the very early stages of the diffusion of
a new technology that is the Web as well until its maturity. The former
is a key point in the lifecycle of a new technology. Firstly, during
this period new technologies are expensive, crude and imperfect
\citep{rosenberg1994exploring, wilson201281}. A simple comparison
between Web 1.0 and Web 3.0 applications clearly illustrates this
\citep{tranos2020social}: while a static website facilitates one-way
information dissemination, a platform like \texttt{github} enables
cooperation between users and the creation of new information, meaning,
and even knowledge \citep{faraj2016special, barassi2012does}. During
this period the performance of a new technology is the main attraction
and not the cost to access and use it \citep{wilson2011lessons}. There
is a broader theoretical discussion in the literature about the early
adoption motives. As summarised by \citet{perkins2005international}, on
one hand, epidemic models highlight the role of interpersonal contacts
as a way for new technologies to diffuse. On the other hand, economic
models underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories \citep{neffke2011regions, simmie2014new}. This
argument is even more relevant when the focus is on digital technologies
because of the commonly found lag between investment and economic
returns as reflected in the Solow paradox
\citep{acemoglu2014return, brynjolfsson2018artificial}.

The data used here depict the active engagement with the Web as they
contain geolocated and time-stamped counts of commercial websites from
the UK. Instead of adopting metrics reflecting the passive engagement
with the internet such as internet adoption and speeds
\citep[e.g.][]{blank2018local, destefano2022fuzzy}, this paper observes
yearly counts of geolocated websites, which reflect the creation and
maintenance of websites. It needs to be highlighted here that although
there are various types of underpinning web technologies -- namely
markup, stylesheet, client- and server-side coding languages as well as
different web browsers and other related programming language -- the
focus here is on the diffusion of the Web in its totality. For a more
detailed and cross-country technological diffusion study, see
\citet{PAPAGIANNIDIS2015308}.

To analyse these data, a diverse set of methods are employed. Firstly,
cumulative adoption curves are estimated for the UK as a whole and for
the Local Authority Districts (LAD), illustrating, for the first time,
such S-shaped diffusion patterns for local areas. This system level
analysis also reveals insights about the volatility of the adoption of
the Web. Then, exploratory spatial data analysis offers some first
indications of the importance of the different spatial diffusion
mechanisms in shaping the diffusion of the Web in the UK. Importantly,
the analysis adopts two distinct geographical scales: the LAD and the
much more detailed scale of the smallest census areas in the UK, the
Output Areas. Lastly, a novel methodological framework based on
well-established Machine Learning (ML) algorithms is developed to test
the role of such mechanisms at these two scales. The results reveal for
the first time the importance and relevance of different spatial
diffusion mechanisms for the diffusion of a tangible, digital technology
at very granular geographical scales.

The structure of the paper goes as follows. Section~\ref{sec-litreview}
reviews the key concepts in the diffusion literature as well as relevant
empirical studies. Then, Section~\ref{sec-datamethods} describes the
data and methods used in this paper. The results of the different stages
of the analysis are presented in Section~\ref{sec-results}. The
discussion of the results and the conlcusions of the paper are offered
in Section~\ref{sec-conclusions}.

\textbf{SAY SOMETHING ABOUT DIST PAPER BY ARP, NOT SPECIFIC EXPLANATORY
VARIABLES}

\section{Literature review}\label{sec-litreview}

Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be
identified. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure \citep{hivner2003facilitating}. On
the other hand, there is a hierarchical processes. Instead of a
horizontal distance-based diffusion mechanism, the top-down hierarchy of
urban systems shapes technological diffusion. In reality, the synthesis
of these two processes better reflects how new technologies diffuse over
space and time \citep{morrill2020spatial}.

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled `Innovation Diffusion as a Spatial Process'
\citep{hagerstrand1968innovation}. Hägerstrand was the first one to
identify diffusion as a geographical process. The starting point was the
idea that diffusion is based on passing information through social
networks, which themselves tend to be defined by geography. Hence, he
identified a `neighbourhood' mechanism of how information, and
consequently, innovation diffuse. He used agricultural innovations to
test and model his ideas using Monte Carlo simulations. Hägerstrand also
incorporated the role of hierarchy and how some phenomena maybe firstly
adopted in larger cities and then diffuse to second tier ones. This is a
sequential instead of a simultaneous process, which resembles the
`lead-lag' spatial acceleration effect in market research
\citep{bento2018time, PERES201091}. Hägerstrand is more widely known
though for highlighting the role time plays in the diffusion of
innovations: an early-pioneering period, a middle fast accelerating
period and a final saturation one \citep{morrill2020spatial}.

The temporal dimension was further explored by Everett Rogers and his
seminal work on `Diffusion of Innovations' \citep{rogers2010diffusion}.
Rogers being a sociologist focused not on the diffusion of innovations
over space and time, but instead on the adoption of new technologies and
innovations by individuals and the individual mechanisms that drive the
decisions behind adoption. He identified five groups of individuals
regarding their adoption speed: innovators, early adopters, early
majority, late majority and laggards. The key mechanism of diffusion and
adoption is communication and how knowledge is transferred within a
social system. All above approaches agree that diffusion and the
cumulative adoption of a new technology are represented by an S-shaped
line plot \citep{grubler1990rise}.

Schmidt's Law empirically illustrates a similar pattern. \emph{Core} and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed \citep{grubler1990rise}. This is
where the first adopters tend to be based. Then, technologies spread to
the \emph{rim} and eventually to the \emph{periphery}. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions \citep{leibowicz2016representing}.

\citet{grubler1990rise} effectively summarises the three key spatial
diffusion mechanisms: (i) the cumulative level of adoption follows an
S-shaped pattern just like purely temporal models; (ii) diffusion is
shaped by a hierarchy mechanisms in a form of a centrifugal force: from
core to periphery; and (iii) diffusion is also shaped by distance and
proximity. These are the three mechanisms that the empirical analysis in
Section~\ref{sec-results} investigates.

The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted in this
paper. \citet{beardsell1999spatial} studied the evolution of the
computer industry in 317 US metro areas during the 1977-1992 period
using employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, \citet{bednarz2020pulled} focused on wind turbines and modeled
their spatial diffusion across 402 German regions during 1970-2015.
Their key finding is that local demand than local supply was the main
driving factor. \citet{haller2011determinants} employed firm panel data
for Ireland during 2003-2005 and illustrated that when firms are located
in a region or industry with high share of firms having a website, they
are more likely to have a website too.

At a global scale \citet{perkins2005international} explored whether the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
\citet{perkins2011internet} explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones were
shaped by similar socioeconomic factors as the internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, \citet{ding2010modeling} modeled the spatial diffusion
of mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network,
\citet{lengyel2020role} analysed its adoption and the churn at a very
granular spatial level. Their results are in agreement with early
theoretical and empirical contributions reviewed here: assortativity,
urban scaling and distance are the key drivers of spatial diffusion. At
a global scale \citet{PAPAGIANNIDIS2015308} modeled the diffusion of
different web technologies and practices across countries.
Interestingly, they used similar, but less extensive data as the ones
used here. Their analysis illustrated how the diffusion of different web
technologies and practices follow an S-shaped pattern as well as the
different diffusion rates of the different technologies and practices.

All in all, this paper tests whether the three distinct spatial
diffusion mechanisms shaped the diffusion of the Web in the UK at very
granular spatial scales from the Web's early stages onwards.

\section{Data and Methods}\label{sec-datamethods}

To capture the diffusion of the Web, a website density metric is
developed for two different geographical scales: the Local Authority
Districts (LAD) and the Output Areas (OA). The former is an
administrative unit and there are 374 LAD in the UK. The latter is the
smallest census-based geographical unit and there are c.~230,000 of them
in the UK. This methodological choice will allow the mapping of the
diffusion of web technologies and the assessment of the diffusion
mechanisms at these two very different spatial scales.

The counts of websites at these scales are calculated using data from
the Internet Archive\footnote{\href{See\%20https://archive.org/}{https://archive.org/}.}
and, specifically, the JISC UK Web Domain Dataset \citep{ukwebarchive}.
The Internet Archive is one of the most complete and oldest archive of
webpages in the world operating since 1996
\citep{ainsworth2011much, holzmann2016dawn}. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage it
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK ccTLD (.uk) from the
1996--2012 period. In essence, this is a long list of 2.5 billion URLs
of archived webpages including also the archival timestamp.

Instead of using the whole .uk ccTLD, this paper focuses on its
commercial subset, the .co.uk second level domain (SLD). This choices
decrease the heterogeneity of the web data as such commercial websites
have specific aims: they are used to diffuse information, support online
transactions and share opinions
\citep{THELWALL2000441, blazquez2018big}. Although a UK company can
adopt a generic TLD such as .com and these cases escape the data used
here, such omissions should not affect the validity of the results given
the popularity of the .uk ccTLD \citep{tranosuk}: UK consumers prefer to
visit a .uk website when they are searching for products or services
\citep{hope}; and canecdotal evidence indicates that during the first
half of 2000, threec .co.uk domains were registered every minute
\citep{oecd_coms}. Importantly, previous studies illustrated that .co.uk
is the most popular UK SLD \citep{tranosuk}.

The text from these webpages was scanned using a regular expression
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode \citep{BL2013geo}. This
information allows the geolocation of the data and the creation of the
LAD and OA counts.

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This
website reconstruction process lend itself to the consteruction of
\emph{website} instead of \emph{webpage} density metrics. Websites tend
to represent specific orgnanisations or entities, and, arguably, are
more meaningful observational units than webpages, which can ignore the
upstream dependency of the website they belong to. Based on the
following example, all three webpages are part of the same website
(\url{http://www.website.co.uk}). Only the overall website and not the
nested webpages are considered for the counts as otherwise the density
metrics would have been biased towards large, place specific, websites.

\begin{itemize}
\item
  \url{http://www.website.co.uk/webpage_a} B15 2TT
\item
  \url{http://www.website.co.uk/webpage_b} BS8 1TH
\item
  \url{http://www.website.co.uk/webpage_c} B15 2TT
\end{itemize}

\noindent What is challenging is that this aggregation approach, which
has been used elsewhere \citep{tranosuk, shoreditch} may lead to
websites with multiple postcodes. As per the above example,
\href{http://www.website.co.uk}{www.website.co.uk} includes two unique
postcodes: B15 2TT and BS8 1SS. The distribution of the number of
postcodes per website for 2000 is presented in Table \ref{f2000}, which
clearly illustrates a wide range. At the left end of the distribution,
there are websites anchored to a unique location (72\% of all the
reconstructed websites in 2000), which may represent a small company
with a single trading location. At the right end, there are websites
with thousands of different postcodes. Considering the time period of
the analysis, such cases can represent directories which used to be
popular in the pre-search engines early times of the commercial internet
as well as real estate websites \citep{tranosuk}.

The analysis presented here is based on two subsets of these data.
Firstly, on websites, which only contain one unique postcode. As a
robustness check, the analysis is replicated for an extended subset of
websites, which include up to 10 unique postcodes to capture commercial
websites which point to multiple locations. They are geolocated by
equally attaching them to these locations. This extended sample includes
94 percent of all the archived websites in 2000.

@@

Another data cleaning step deals with some extreme outliers. Figure
\ref{correct} plots the website counts for the top 1000 postcodes. Some
obvious outliers can be observed for the 2002-2006 period for a handful
of postcodes, which can be attributed to link farms \citep{BL2013links}.
The website counts for these postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1
6GL, M28 2SL, DE21 7BF), which in 2004 or 2005 had more than 1000
websites pointing to them, were replaced with predicted values based on
a simple panel regression model with postcode fixed effects and yearly
dummy variables. These postcodes refer to a residential area with a
small park in South London, residential areas outside Warwick and
Manchester, a rural area in the south border of England and Wales and
small business parks in Crew and Derby. Hence, it is difficult to
believe that these postcodes genuinely hosted such a large amounts of
websites, which were later extensively declined. To put the magnitude of
the data imputation into perspective, this process affected 6 out of the
557,808 postcodes included in the data. As the Section~\ref{sec-results}
illustrates, this led to a small increase of the model predictive
capability.
